Retrofitting Effect Handlers onto OCaml

### 2020-11-09 retro-concurrency.tex
Updates

What makes effect handlers quite powerful is that they provide a structured
form of delimited continuations~\cite{}, which enables advanced non-local
control-flow mechanisms such as resumable exceptions, coroutines, generators
and asynchronous I/O to be composably expressed. As an example, consider the
task of handling division-by-zero case in a calculator application. Different
calculators handle division-by-zero differently. In the iOS 13 Calculator App
(|iOS|), division-by-zero returns |Error|. In the Google search calculator App
(|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
|n/0| returns |Infinity| with the same sign as |n|. How can we capture these
different behaviours without changing the evaluator?

Common Lisp Conditions~\cite{Conditions} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision. If
the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We can implement
this mechanism using effect handlers as shown below:

@@

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient data and task parallel programs in the vein
of Cilk~\cite{}, OpenMP~\cite{} and Intel TBB~\cite{}.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

>>>>

What makes effect handlers quite powerful is that they provide a structured
form of delimited continuations~\cite{Danvy90}, which enables advanced
non-local control-flow mechanisms such as resumable exceptions, coroutines,
generators and asynchronous I/O to be composably expressed. As an example,
consider the task of handling division-by-zero case in a calculator
application. Different calculators handle division-by-zero differently. In the
iOS 13 Calculator App (|iOS|), division-by-zero returns |Error|. In the Google
search calculator App (|gsearch|) |0/0| returns to |Error|, and for any other
|n| not equal to |0|, |n/0| returns |Infinity| with the same sign as |n|. How
can we capture these different behaviours without changing the evaluator?

Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision. If
the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We can implement
this mechanism using effect handlers as shown below:

@@

There are many research languages and libraries built around effect
handlers~\cite{Koka,Links,Pyro,Frank,Eff}. Unlike these efforts, our goal is to
retrofit effect handlers onto the OCaml programming language, which has been in
continuous use for the past 25 years in large codebases including verification
tools~\cite{FStar,Coq}, mission critical software systems~\cite{astree} and
latency sensitive networked applications~\cite{JS,Docker,MirageOS}. OCaml is
particularly favoured for its competitive yet predictable performance, with
fast foreign-function interface (FFI) and excellent compatibility with program
analysis tools such as debuggers and profilers that utilise DWARF stack unwind
tables~\cite{DWARF} to obtain a backtrace. Our primary motivation for adding
effect handlers to OCaml is to support (1) highly concurrent applications such
as web servers to be expressed naturally in \emph{direct-style} as opposed to
using callbacks and (2) efficient nested parallel programs in the vein of
Cilk~\cite{Cilk}, OpenMP~\cite{}, Habenero Java~\cite{Imam14}, X10~\cite{},
Intel TBB~\cite{IntelTBB}, parallel Haskell variants~\cite{} and parallel ML
variants~\cite{}.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{retro-concurrency}
